---
title: "STATS 769 - Lab 05 - bole001"
author: "Bernard O'Leary"
date: "02 September 2019"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data format and API call details

Data is in JSON format as it comes down from the API. It is a JSON array of "trip" objects. The API allows query-sting based parameters to be passed as part of an HTTP GET request, which means that we don't need to send an HTTP body. The paramters that we use are *limit* to tell the API to give us back a specific number of JSON rows. Without this parameter we get back 1000 by default. We also send a *year* and a *vehicle_type* paramter to further limit our result set. Finally an *api_token* parameter is sent to authenticate our request. I am using a combination of the the *httr* and *jsonlite* libraries to make a simple GET request and then transform the result into a dataframe.

The following code imports the data from the API and lists dataframe dimensions and first 6 rows of the data for validation purposes.

```{r}
library(httr)
library(jsonlite)

# Call the resource
json_result <- GET("https://data.austintexas.gov/resource/7d8e-dm7r.json?$limit=10000&year=2018&vehicle_type=scooter&$$app_token=xo9XRO6BRv2CwPMK9RfhyFzN6")

# Show data frame
trips <- fromJSON(content(json_result, as="text"))
head(trips)
dim(trips)

```
## Cleanse dataset

Subset only trips with non-negative distances and durations, create a new *long_trip* variable (where "long" means that the trip distance was greater than 1000m). We exclude non-positive durations and distances and make another new variable that is a log of the duration variable called *logged_trip_duration*. Show the first six rescords of the dataframe so that we can see these new variables and validate. 

```{r}
model_trips <- subset(trips, as.integer(trip_duration) > 0 & as.integer(trip_distance) > 0)
model_trips$logged_trip_duration <- log(as.integer(model_trips$trip_duration))
model_trips$long_trip <- as.integer(model_trips$trip_distance) > 1000
head(model_trips)
```

## Boxplot the data (to see what we have)

Take a look at the data that we have come back from the API call. We have at least one significant outlier. Other than that looks like we have perhaps a 

```{r}
plot(log(as.integer(model_trips$trip_distance)) ~ model_trips$long_trip, ylab="distance", xlab="long trip")
barplot(table(model_trips$long_trip), main="Long Trip (TRUE/FALSE)")
```

# Comparison of logistic regression and k-nearest neighbours models

We fit a logistic model that predicts the proportion of long trips as a function of trip duration and output the confusion matrix. We do that same for KNN and then try to plot a chat of the result. Unfortunately I have not been able to produce a chart successfully.

```{r}
test_index <- sample(nrow(model_trips), nrow(model_trips) * 0.1)
test_trips <- model_trips[test_index, ]
train_trips <- model_trips[-test_index, ]

overall_proportion <- mean(model_trips$long_trip)
glm_fit <- glm(y ~ x, data.frame(x=train_trips$logged_trip_duration, y=train_trips$long_trip), family="binomial")

library(caret)
glm_prob <- predict(glm_fit, data.frame(x=as.integer(test_trips$trip_distance)), type="response")
glm_pred <- ifelse(glm_prob > .5, 1, 0)
glm_diag <- confusionMatrix(factor(glm_pred, levels = 1:0), 
                            factor(as.integer(test_trips$long_trip), levels = 1:0))
glm_diag$table
glm_diag$overall["Accuracy"]
```

```{r}
train_x <- as.data.frame(train_trips$logged_trip_duration)
test_x <- as.data.frame(test_trips$logged_trip_duration)
train_y <- as.data.frame(as.integer(train_trips$long_trip))
cl = train_y[,1, drop = TRUE]

library(class)
knn_pred <- knn(train_x, test_x, cl, k=31, prob=TRUE)
knn_prob <- ifelse(knn_pred == 1, attr(knn_pred, "prob"), 1 - attr(knn_pred, "prob"))
knn_diag <- confusionMatrix(factor(knn_pred, levels = 1:0), 
                            factor(as.integer(test_trips$long_trip), levels = 1:0))
knn_diag$table

knn_diag$overall["Accuracy"]
par(mar=c(3, 3, 2, 2))
breaks <- seq(0, 1, .1)
midbreaks <- breaks[-1] - diff(breaks)/2
class(test_x)
props <- tapply(as.integer(test_trips$long_trip), cut(test_trips$logged_trip_duration, breaks), mean)
props
midbreaks
plot(midbreaks, props, pch=16, ylim = c(0,1))
lines(test_trips$logged_trip_duration, glm_prob, col="red")
lines(test_trips$logged_trip_duration, knn_prob, col="green")

```

# Summary of analysis

The result of the KNN model is better than for the logistic regression, at least for the Accuracy metric. For logistic regression we have Accuracy of about 51% and for KNN we have accuracy of over 80%. For this dataset and scenario at least KNN is a better model than loogistic regression in terms of Accuracy.
