
# 769 Lab Report Template

This document provides a (very) rough exemplar of the sort of report
that should be submitted for each lab.  The report should NOT exceed
10 pages (when printed at 100% scale).

## Data Import 

The data set consists of two large CSV files (~500MB) containing
information about flights in the United States.
These files are too large
for us to deal with easily (until we cover Large Data later in the course),
so we just worked with a sample of 10,000 rows from each file.

We obtained those samples using Linux shell tools (because doing the 
job in R would take too long and use too much memory).  We will learn
about these tools in the "Linux Shell" topic.

```{r}
system("head -1 /course/ASADataExpo2009/Data/1988.csv | tee 1988.csv > 1989.csv")
system("shuf -n10000 /course/ASADataExpo2009/Data/1988.csv >> 1988.csv")
system("shuf -n10000 /course/ASADataExpo2009/Data/1989.csv >> 1989.csv")
```

Once we have the samples, reading CSV files into R is very easy.

```{r}
flights1988 <- read.csv("1988.csv")
flights1989 <- read.csv("1989.csv")
```

It is also easy to combine the resulting data frames into a single
data frame.

```{r}
flights <- rbind(flights1988, flights1989)
dim(flights)
```

For most topics in the first half of the course, this Data Import section
will be the largest part of the report (so the "Analysis" sections
will have to be quite brief).

## Exploratory Analysis

It will usually be useful to calculate some basic summaries and
draw some basic plots of the data set, especially the first time
we work with a data set.

This table and plot show that the number of flights per month is quite even.

```{r}
table(flights$Month)
barplot(table(flights$Month))
```

This plot shows a weak suggestion that (the log of)
departure delays are greater 
later in the day.

```{r}
plot(flights$DepTime, flights$DepDelay, log="y")

```

## Modelling

We will (usually) build a predictive model for each lab, though the
quality of the model will not (usually) be the focus of the lab.
Lab work will be more focused on importing and tidying and transforming
the data, and dealing with issue like data size and speed of
computation.

We will (usually) work with a training set of data to build a model
and then assess the model on a test set.  Here we use 1988 data
for training and 1989 data for testing.

Modelling will often require further transformation and restructuring
of the data.  In this case, we log the departure delays, which means
we must discard all non-positive delays.

```{r}
trainDelay <- log(flights1988$DepDelay)
trainSubset <- is.finite(trainDelay)
yTrain <- trainDelay[trainSubset]
xTrain <- flights1988$DepTime[trainSubset]
```

We consider two models:  a simple constant mean delay and a linear
regression of delay against departure time.

```{r}
fitMean <- mean(yTrain, na.rm=TRUE)
fitLM <- lm(y ~ x, data.frame(y=yTrain, x=xTrain))
```

It may be useful to draw plots to show the fitted models.
In this case, there is only a quite weak signal and lots of
unexplained variability.

```{r}
plot(xTrain, yTrain)
abline(h=fitMean, col="red", lwd=3)
abline(fitLM, col="blue", lwd=3)
```

We assess the model on the test set;  because this is a linear regression
we calculate the Root Mean Square Error.  The test set must be
logged to correspond to the training set.

```{r}
testDelay <- log(flights1989$DepDelay)
testSubset <- is.finite(testDelay)
yTest <- testDelay[testSubset]
xTest <- flights1989$DepTime[testSubset]
predMean <- rep(fitMean, length(yTest))
predLM <- predict(fitLM, data.frame(x=xTest))
```
The linear regression is a slightly better predictor than a simple
constant mean.

```{r}
RMSE <- function(m, o) {
    sqrt(mean((m - o)^2))
}
RMSE(predMean, yTest)
RMSE(predLM, yTest)
```

Again, a plot may help to show the performance of the models on the 
test set.

```{r}
plot(xTest, yTest)
abline(h=predMean, col="red", lwd=3)
abline(fitLM, col="blue", lwd=3)
```

## Conclusion

We will (usually) want to summarise the results of our modelling.
In this case, we have a pretty poor model, but we have still done better
than just taking a simple average.

We will also (usually) want to reflect (briefly) on the course topic
that the lab is built around (e.g., Linux shell tools);  what have
we used, why, and how successful were we?
